---
title: "Homework Assignment 4"
subtitle: "Prioritizing potential aquaculture"
author: "IMS"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-fold: true
#| code-summary: "Load packages"
library(terra)
library(sf)
library(tidyverse)
library(here)
library(tmap)
library(kableExtra)
```


## Prepare Data

Load West coast shapefile and bathymetry raster.
```{r}
WC_EEZ <- sf::read_sf(here("data", "wc_regions_clean.shp"))

bath <- terra::rast(here("data", "depth.tif"))
```

Load Rasters and stack.

```{r}
SST_files <- list.files(
  here("data"), pattern = "average", 
  full.names = TRUE)

SST_stack <- terra::rast(SST_files)
names(SST_stack) <- c("2008", "2009", "2010", "2011", "2012")
```

```{r}
#| code-fold: true
#| code-summary: "Check if CRS of each dataset are equal, update if not"

if(sf::st_crs(WC_EEZ) == sf::st_crs(4326)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  WC_EEZ <- sf::st_transform(WC_EEZ, st_crs(4326))
}

if(terra::crs(bath) == terra::crs("EPSG:4326")) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  bath <- project(bath, crs("EPSG:4326"))
}

if(terra::crs(SST_stack) == terra::crs("EPSG:4326")) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  SST_stack <- terra::project(SST_stack, crs("EPSG:4326"))
}
```


## Process data

###Next, we need to process the SST and depth data so that they can be combined. In this case the SST and depth data have slightly different resolutions, extents, and positions.

Find the mean SST from 2008-2012 (e.g. create single raster of average SST)
```{r}
SST_mean_K <- terra::mean(SST_stack)
```

convert average SST from Kelvin to Celsius
hint: subtract by 273.15

```{r}
SST_mean_C <- SST_mean_K - 273.15
```

crop depth raster to match the extent of the SST raster

```{r}
bath <- terra::crop(bath, SST_mean_C)
```

note: the resolutions of the SST and depth data do not match
resample the depth data to match the resolution of the SST data using the nearest neighbor approach

```{r}
bath <- terra::resample(bath, SST_mean_C, "near")
```


check that the depth and SST match in resolution, extent, and coordinate reference system
hint: can the rasters be stacked?

```{r}
if(all(terra::res(bath) == terra::res(SST_mean_C))) {
  if(terra::crs(bath) == terra::crs(SST_mean_C)) {
    if(terra::ext(bath) == terra::ext(SST_mean_C)){
      print("All match")
    }else{
      stop("extenet does not match")
    }
  }else{
    stop("CRS does not match")
  }
}else{
  stop("Resolution does not match")
}
```

```{r}
SST_bath_test <- terra::rast(c(SST_mean_C, bath))
```

## Find suitable locations

reclassify SST and depth data into locations that are suitable for oysters
hint: set suitable values to 1 and unsuitable values to 0

sea surface temperature: 11-30°C
depth: 0-70 meters below sea level

```{r}
# check values before rcl
summary(SST_mean_C)
summary(bath)
```


```{r}
# reclassify erroneous values as NA
SST_rcl <- matrix(c(-Inf, 11, NA,
                 11, 30, 1,
                30, Inf, NA), ncol = 3, byrow = TRUE)

bath_rcl <- matrix(c(-Inf, -70, NA,
                 -70, 0, 1,
                0, Inf, NA), ncol = 3, byrow = TRUE)

SST_mean_C_rcl <- classify(SST_mean_C, rcl = SST_rcl)
bath_rcl <- classify(bath, rcl = bath_rcl)

# check values are 1 or NA
unique(SST_mean_C_rcl)
unique(bath_rcl)
```

```{r}
# Perform the operation: Both cells equal to 1
suitable_loc <- (SST_mean_C_rcl*bath_rcl)
plot(suitable_loc)
```

# Determine the most suitable EEZ

We want to determine the total suitable area within each EEZ in order to rank zones by priority. To do so, we need to find the total area of suitable locations within each EEZ.

select suitable cells within West Coast EEZs
```{r}
EEZ_raster <- terra::rasterize(WC_EEZ, suitable_loc, field="rgn_id")
EEZ_raster <- (EEZ_raster*suitable_loc)
plot(EEZ_raster)
```

find area of grid cells
```{r}
# Calculate the area of each cell (in square meters)
cell_area <- terra::cellSize(EEZ_raster, unit = "km")
```

find the total suitable area within each EEZ
hint: it might be helpful to rasterize the EEZ

```{r}
zone_area <- terra::zonal(cell_area, EEZ_raster, fun="sum", unit="km")
zone_area
```

```{r}
# To map our eez area with geometry later on
EEZ_join <- left_join(WC_EEZ, zone_area, by = "rgn_id")
```

```{r}
kable(EEZ_join[c("rgn", "area")] %>% st_drop_geometry(), digits = 2,
      caption = "Suitable area by exclusive economic zone", 
      col.names = c("EEZ Region", "Area (km^2)"))
```


```{r}
tmap_mode("plot")
tm_shape(bath) +
  tm_raster(palette = "-RdBu",
            title = "Bathymetry\n(m above and below sea level)",
            alpha = 0.7,
            midpoint = 0,
            legend.show = TRUE) +
  tm_shape(EEZ_join, raster.downsample = TRUE) +
  tm_polygons(col = "area",
              palette = "Greens",
              alpha = 0.7,
              title = expression("Suitable habitat area (km"^2*")")) +
  tm_text("rgn", size = 0.4) +
  tm_compass(size = 2,
             position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(legend.outside = TRUE,
            frame = FALSE,
            main.title = "Oyster Suitable Habitat in\nWest Coast Exclusive Economic Zones",
            main.title.size = 1.35,
            main.title.position = c("center", "top"))
```

